---
title: "scNMTseq challenge analysis using a PLS-based approach"
author: Al JalalAbadi^[Melbourne Integrative Genomics, The University of Melbourne (al.jal.abadi@gmail.com)]
date: "`r format(Sys.Date(), '%d %b %Y')`"
output: rmarkdown::html_vignette
params:
  drop_lineages: !r c('Visceral_endoderm')
  keep_assays: !r c("rna", "met_genebody", "met_promoter", "met_cgi", "met_p300", "met_CTCF", "met_DHS")
  save_output: no
  on_my_mac: !r Sys.info()['user'] == 'alabadi'
  not_on_my_mac: !r Sys.info()['user'] != 'alabadi'
vignette: >
  %\VignetteIndexEntry{scNMTseq challenge analysis using a PLS-based approach}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

To Do

* Add the number of features to subtitles in `cov_plot`
* Ensure python3 can be configured for MOFA2

--------

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

**Note:** These analyses are (slightly) different from the ones presented in the hackathon in the sense that:

  - The cells used are different (only embryonic cells used)
  - To enhance reproducibility on different machines, `uwot::umap` was used instead of `umap::umap` for dimensionality reduction of RNA data so that python depndencies are reduced
  
--------

Load the required packages:

```{r, warning=FALSE, message=FALSE}
library(BIRSBIO2020.scNMTseq.PLS)
library(MultiAssayExperiment)
library(ggplot2)
library(knitr)
library(data.table)
library(uwot)
library(reticulate)
```

```{r}
## you might need to set up your own python3 using reticulate::use_python
py.res <- tryCatch({use_python('usr/bin/python3', required = TRUE)}, error = function(e) e)

message('Python used: ', paste0(system('which python', intern = TRUE)))
message(paste0(py.res,'\n', paste(reticulate::py_config(), collapse = '\n')))
```

# Data

Load MAE object from https://cloudstor.aarnet.edu.au/ 

```{r, eval=params$on_my_mac, include=FALSE}
gastru.mae <- readRDS('/Users/alabadi/Projects/analysis/R/multiOmics/scNMTseq/BIRSBIO_scNMTseq/output/scnmtseq_gastrulation_mae_826-cells_orderedFeatures-weights.rds')
```

## filter extra-embryonic cells

```{r, eval=params$not_on_my_mac}
scnmtseq_url <- "https://cloudstor.aarnet.edu.au/plus/s/Xzf5vCgAEUVgbfQ/download?path=%2Foutput&files=scnmtseq_gastrulation_mae_826-cells_orderedFeatures-weights.rds"
gastru.mae <- readRDS(url(scnmtseq_url))
```

```{r}
drop_lineages <- c('ExE_ectoderm', 'Primitive_endoderm', 'Visceral_endoderm')
drop_cells <- (gastru.mae$lineage10x_2 %in% drop_lineages) | is.na(gastru.mae$lineage10x_2)
gastru.mae <- gastru.mae[, !drop_cells,]
```

```{r}
table(gastru.mae$stage) %>% as.data.frame() %>% set_colnames(c('stage', '# of cells')) %>% kable()
```

Check column names are consistent:

```{r}
all_identical(colnames(gastru.mae))
```

```{r}
coldata <- data.frame(colData(gastru.mae))
```

Breakdown of the number of cells in each stage and putative lineage:
```{r, out.width='80%'}
table(coldata$lineage10x_2, coldata$stage) %>% kable
```

## Feature detection

Create density plots of the feature detection rate across all cells for all modalities:

```{r}
# get the methylation assays
met_assays <- grep(names(gastru.mae), pattern = '^met', value = TRUE)
# calculate the feature detection in a data.frame for methylation assays
coverages <- lapply(named_list(met_assays), function(assay_name) {
  mat <- assay(gastru.mae, assay_name)
  NAs <- rowSums(!is.na(mat))/dim(mat)[2]*100
  data.frame(pct_NAs=NAs)
})
# create a long data.frame containing the assay name for plot
coverages <- rbindListWithNames(coverages)
coverages$dataset <- factor(coverages$dataset, levels = unique(coverages$dataset), ordered = TRUE)
```

```{r, fig.width=8, fig.asp=0.4, fig.cap="density plots for methylation data show that shorter genomic regions tend to have less feture coverage. Dashed blue line indicates the average across all modalities."}
cov_plot <- ggplot(coverages, aes(x = pct_NAs)) + geom_density(fill = 'lightblue', show.legend = FALSE) + 
  geom_vline(aes(xintercept=mean(pct_NAs)),
             color="blue", linetype="dashed", size=0.5) +
  labs(x = '% of cells detecting the feature') + facet_wrap(.~dataset, nrow = 2) + 
  theme_bw() + theme(strip.text.x = element_text(size = 10, face = 'bold'))
cov_plot
```

```{r, eval=FALSE, echo=FALSE}
ggsave(cov_plot, filename = 'figures/covplots.pdf', width = 8, height = 4)
```

# RNA

```{r}
rna <- assay(gastru.mae, 'rna')
```

## UMAP

```{r}
ncomp <- 5
set.seed(42)
## all params use defaults except for 'n_components' and 'min_dist'
umap_rna_unsup <- uwot::umap(
  t(rna), 
  n_neighbors = 15,
  n_components = ncomp,
  metric = "euclidean",
  n_epochs = NULL,
  learning_rate = 1,
  scale = FALSE,
  init = "spectral",
  init_sdev = NULL,
  spread = 1,
  min_dist = 0.7, ## as we're interested in continuous transitions rather than 'clusters'
  set_op_mix_ratio = 1,
  local_connectivity = 1)
```

```{r, eval=FALSE, echo=FALSE}
saveRDS(umap_rna_unsup, file = sprintf('savedata/umap-result-ncomp-%s.rds', ncomp))
```

Get a data.frame of embeddings:
```{r}
df <- data.frame(umap_rna_unsup)
colnames(df) <- paste0('UMAP_', seq_along(df))
df <- cbind(df, coldata)
```

Create colour palettes for stages and lineages:

```{r}
nstage <- length(unique(gastru.mae$stage))
nlineage <- length(unique(gastru.mae$lineage10x_2))
ntotal <- nstage + nlineage
col_pallete <- gg_color_hue(n=ntotal)
ind.stage <- 2*seq_len(nstage) ## every other colour
stage_cols <- col_pallete[ind.stage]
lineage_cols <- col_pallete[-ind.stage]
```


```{r}
## helper function to create both plots coloured by stage and lineage
plot_umap_by_stage_and_lineage <- function(df, dims = c(1,2)) {
  axes <- paste0('UMAP_', dims)
  p1 <- ggplot(df, aes_string(axes[1], axes[2])) + geom_point(aes(col=lineage10x_2)) + 
    theme_classic() + scale_color_manual(values = lineage_cols)
  
  p2 <- ggplot(df, aes_string(axes[1], axes[2])) + geom_point(aes(col=stage)) + 
    theme_classic()+ scale_color_manual(values = stage_cols)
  list(lineage = p1, stage = p2)
}
```

First two components highlight the distinct transcriptional state of the early-stage cells compared to the late-stage cells:

```{r, fig.asp=0.6}
plot_umap_by_stage_and_lineage(df = df, dims = c(1,2))
```

The variation driven by known and putative phenotypes is more apparent in the following dimensions:

```{r, fig.asp=0.6}
plot_umap_by_stage_and_lineage(df = df, dims = c(2,3))
```

```{r, fig.asp=0.6}
plot_umap_by_stage_and_lineage(df = df, dims = c(3,4))
```

